---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dpi = 300,
  out.width = "100%"
)
```

# PaleoSpec

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/EarthSystemDiagnostics/paleospec/branch/master/graph/badge.svg)](https://app.codecov.io/gh/EarthSystemDiagnostics/paleospec?branch=master)
<!-- badges: end -->

PaleoSpec is an R package to assist in the spectral analysis of timeseries, in particular timeseries of climate variables from observational, model, and proxy paleoclimate data sources. PaleoSpec contains functions to analyse existing timeseries and to generate timeseries with specific spectral properties.


## Installation

You can install the development version of PaleoSpec from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("EarthSystemDiagnostics/paleospec")
```


Please refer to function references here:

https://earthsystemdiagnostics.github.io/paleospec/reference/index.html



## Usage

### Simulating timeseries with given spectral properties

`SimPLS` can be used to create a timeseries whose power spectrum has powerlaw like properties, 
where: $S(f) = \alpha f^{-\beta}$ 

```{r}
library(PaleoSpec)
set.seed(20221109)
N <- 1e04

alpha <- 0.1
beta <- 1

ts1 <- SimPLS(N = N, b = beta, a = alpha)

plot(ts1, type = "l")
```

`SpecMTM` can be used to estimate the power spectrum of a timeseries using the multitaper method (ref)

```{r}
sp1 <- SpecMTM(ts1)
```

The `LPlot` function will plot a power spectrum on log10-log10 axes.

```{r}
LPlot(sp1)
abline(log10(alpha), -beta, col = "red")
```

You can add confidence intervals to the spectral estimates with `AddConfInterval`

```{r}
sp1 <- AddConfInterval(sp1)
LPlot(sp1)
abline(log10(alpha), -beta, col = "red")
```


The `LogSmooth` function can be used to smooth power spectra with equally spaced filter in log-space.

```{r}
sp1_f <- LogSmooth(sp1, df.log = 0.01)
LPlot(sp1)
LLines(sp1_f, col = "green")
```



