---
title: "Comparison of SpecACFBiasFree vs SpecACF"
format:
  html:
    toc: true
    code-fold: true
    fig-width: 10
    fig-height: 7
execute:
  cache: false
---

Three scenarios × three estimators evaluated by Monte Carlo:

**Scenarios**

| Label | Description |
|-------|-------------|
| none | No gaps (sanity check) |
| random | 25% independent random gaps (Damaschke "outliers" case) |
| extended | 25% in runs (mean run = 8 tu; Damaschke "extended gaps") |

**Estimators**

| Label | Description |
|-------|-------------|
| `SpecACF` | k = 1, untapered comparison point |
| `SpecACFBiasFree_trunc` | Linear ACF, K = N/4 (Damaschke-recommended truncation) |
| `SpecACFBF_trunc_k5` | As above, plus k = 5 DPSS tapers (nw = 3) |

**Diagnostics**

1. **Bias** — mean spectrum vs the true PSD, log-binned
2. **Scatter** — IQR / median across realisations
3. **% negatives** — fraction of frequencies with negative power

## Setup

```{r setup}
#| cache: false
suppressMessages({
  if (requireNamespace("devtools", quietly = TRUE) &&
      file.exists("DESCRIPTION")) {
    devtools::load_all("..", quiet = TRUE)
  } else {
    library(PaleoSpec)
  }
})

set.seed(20240505)
library(ggplot2)
```

## Simulation parameters

```{r params}
N        <- 1000      # series length
beta     <- 0         # power-law slope of true spectrum
alpha    <- 0.1       # power-law amplitude  S(f) = alpha * f^-beta
n_reps   <- 100       # Monte Carlo replications per scenario
gap_frac <- 0.25
gap_run  <- 8         # mean length of extended-gap runs
```

## Gap generators

```{r gap-generators}
gap_none      <- function(n) rep(FALSE, n)
gap_random    <- function(n) runif(n) < gap_frac
gap_extended  <- function(n) {
  # Two-state Markov chain: p(invalid) = gap_frac, mean run = gap_run.
  p_stay_inv   <- 1 - 1 / gap_run
  p_become_inv <- gap_frac * (1 - p_stay_inv) / (1 - gap_frac)
  s <- logical(n); s[1] <- runif(1) < gap_frac
  for (i in 2:n)
    s[i] <- if (s[i - 1]) runif(1) < p_stay_inv else runif(1) < p_become_inv
  s
}

scenarios <- list(
  none     = gap_none,
  random   = gap_random,
  extended = gap_extended
)
```

## Estimators

```{r estimators}
estimators <- list(
  SpecACF               = function(x) SpecACF(x, bin.width = 1, k = 1, nw = 0,
                                              TrimNA = FALSE),
  SpecACFBiasFree_trunc = function(x) SpecACFBiasFree(x, bin.width = 1,
                                                       lag.max = N %/% 4),
  SpecACFBF_trunc_k5    = function(x) SpecACFBiasFree(x, bin.width = 1,
                                                       lag.max = N %/% 4,
                                                       k = 5, nw = 3)
)
```

## Monte Carlo run

```{r monte-carlo}
run_one <- function(gap_fn, est_fn) {
  x <- SimPLS(N, beta, alpha)
  x[gap_fn(N)] <- NA
  est_fn(x)
}

mc <- list()
for (s_name in names(scenarios)) {
  for (e_name in names(estimators)) {
    message("running ", s_name, " / ", e_name, " ...")
    mc[[paste(s_name, e_name, sep = "::")]] <-
      replicate(n_reps, run_one(scenarios[[s_name]], estimators[[e_name]]),
                simplify = FALSE)
  }
}
```

## Helper functions

```{r helpers}
summarise_one <- function(reps) {
  freqs <- reps[[1]]$freq
  Smat  <- vapply(reps, function(r) r$spec, numeric(length(freqs)))

  Smean   <- rowMeans(Smat, na.rm = TRUE)
  Smedian <- apply(Smat, 1, median,   na.rm = TRUE)
  Slo     <- apply(Smat, 1, quantile, probs = 0.25, na.rm = TRUE)
  Shi     <- apply(Smat, 1, quantile, probs = 0.75, na.rm = TRUE)

  p_neg_per <- vapply(reps, function(r) mean(r$spec < 0, na.rm = TRUE),
                      numeric(1))

  list(f = freqs, Smean = Smean, Smedian = Smedian,
       Slo = Slo, Shi = Shi, p_neg = mean(p_neg_per))
}

true_at      <- function(f) alpha * f^-beta

rel_lin_bias <- function(s, est) {
  vals <- (s[[est]] - true_at(s$f)) / true_at(s$f)
  mean(vals[is.finite(vals)])
}
```

## Summarise replicates

```{r summarise}
summary_table <- list()
for (key in names(mc)) {
  parts <- strsplit(key, "::", fixed = TRUE)[[1]]
  summary_table[[key]] <- c(scenario  = parts[1],
                             estimator = parts[2],
                             summarise_one(mc[[key]]))
}
```

## Summary table

Relative bias close to 0 = unbiased. Scatter = mean (Q3 − Q1) / median across
frequencies. % neg = mean fraction of frequencies with negative power.

```{r summary-table}
rows <- lapply(names(summary_table), function(key) {
  s       <- summary_table[[key]]
  scatter <- mean((s$Shi - s$Slo) / abs(s$Smedian), na.rm = TRUE)
  data.frame(scenario  = s$scenario,
             estimator = s$estimator,
             rel.bias  = rel_lin_bias(s, "Smean"),
             scatter   = scatter,
             pct.neg   = round(100 * s$p_neg, 1))
})
tbl <- do.call(rbind, rows)
knitr::kable(tbl, digits = 3,
             col.names = c("Scenario", "Estimator", "Rel. bias", "Scatter", "% neg"))
```

## Plots

```{r plot-prep}
#| include: false
cols <- c(SpecACF               = "red",
          SpecACFBiasFree_trunc = "darkgreen",
          SpecACFBF_trunc_k5    = "purple")

make_df <- function(key, extra_cols) {
  s     <- summary_table[[key]]
  parts <- strsplit(key, "::", fixed = TRUE)[[1]]
  cbind(
    data.frame(scenario  = factor(parts[1], levels = names(scenarios)),
               estimator = factor(parts[2], levels = names(estimators)),
               f         = as.numeric(s$f)),
    extra_cols(s)
  )
}

df_spec <- do.call(rbind, lapply(names(summary_table), make_df,
  extra_cols = function(s) data.frame(Smean = as.numeric(s$Smean))))

df_scatter <- do.call(rbind, lapply(names(summary_table), make_df,
  extra_cols = function(s) data.frame(
    spread = as.numeric(s$Shi) / as.numeric(s$Slo))))
df_scatter <- subset(df_scatter, is.finite(spread) & spread > 0)

ftrue   <- exp(seq(log(min(df_spec$f)), log(0.5), length.out = 200))
df_true <- data.frame(f = ftrue, S = true_at(ftrue))
```

### Mean spectrum vs truth

```{r plot-mean-spec}
#| fig-cap: "Mean spectrum across realisations (solid) vs true power law (dashed). Log–log axes."
ggplot(df_spec, aes(f, Smean, colour = estimator)) +
  geom_line(data = df_true, aes(f, S), inherit.aes = FALSE,
            colour = "black", linetype = "dashed", linewidth = 0.8) +
  geom_line(linewidth = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = cols) +
  facet_grid(estimator~scenario, labeller = label_both) +
  labs(x = "frequency", y = "S(f)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

### Spectral scatter (IQR)

```{r plot-scatter}
#| fig-cap: "Q3 / Q1 ratio across realisations. Closer to 1 = less scatter."
ggplot(df_scatter, aes(f, spread, colour = estimator)) +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(linewidth = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = cols) +
  facet_wrap(~scenario, labeller = label_both) +
  labs(x = "frequency", y = "Q3 / Q1 (spread)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```
